---
title: "BEES3041_CoralDiversity_Bleaching"
author: "Sophie"
date: "7/19/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The class "Anthozoa" has more than 1.5 million data entries. It comprises 11 orders, namely "Actiniaria", "Alcyonacea", "Antipatharia", "Corallimorpharia", "Heliolitina", "Helioporarea", "Penicillaria", "Pennatulacea", "Scleractinia", "Spirularia" & "Zoantharia". However, not all of these orders are tropical corals.
Specifically, "Actiniaria", "Penicillaria" and "Spirularia" are sea anemones, "Pennatulacea" are sea pens, "Antipatharia" are deep sea corals and "Heliolitina"are fossils. Hence, all these 6 orders are neglected for this project. Only the following orders are considered: 
[1]  Alcyonacea: the soft corals
[2]  Corallimorpharia: tropical corals, closely related to Scleractinia
[3]  Helioporacea: colonial, stony corals
[4]  Scleractinia: stony corals & gorgonians
[5]  Zoantharia
The data from the above named orders is downloaded indivdially from gbif.org and saved on to the computer.

```{r Load libraries}
library(data.table) # needed for fread
library(sf)
library(fasterize)
library(dplyr) # included in tidyverse???
library(maps)
library(maptools)
library(sp)
library(raster)
library(CoordinateCleaner)
library(rnaturalearth)
library(tidyverse)
library(ggplot2) # included in tidyverse????
library(viridis) # nice color set
library(plotly) # stacked barplot
library(RColorBrewer)
library(purrr) # map over a function
library(magick) # necessary for animation and reading pngs
library(ggmap) # mapping with ggplot
```

The data that has been successfully saved on to the computer (in form of .csv files) is now loaded into r studio.

```{r Load data}

coral_list <- list.files(pattern="order.*csv")

load_data <- function(data_list) {
   coral_data_list<-list()
   e<-1
   for (l in data_list) {
     coral_data_list[[e]] <- fread(l, select = c("species","decimalLongitude", "decimalLatitude", "speciesKey"))
     e=e+1
   }
   return(coral_data_list)
 }
 
coral_data <- load_data(coral_list)

```

In the different datasets, some records are incomplete; i.e. missing values in either coordinates or species name. This step therefore consists in the removal of N.A.s in coordinates ("decimalLongitude" & "decimalLatitude") and "speciesKey".

Note: if longitude is missing, latitude is missing as well. Hence, no need to check for N.A.s in decimalLatitude. Similarly, if "speciesKey" is not indicated, the species name is not known either.

```{r Data cleaning}
 
 clean_data <- function(df) {
     filt_data <- filter(df,!is.na(decimalLongitude)&!is.na(speciesKey)) %>%
     list();
   } 

filtered_list <- lapply(coral_data,FUN=clean_data)

```

The coorindate records can sometimes be very inaccurate/misleading to the extent that the indicated coordinates for a coral species correspond to a location on land. Given that this project exclusively focuses on marine coral species, all the records with "land" coordinates will be removed via the "clean_coordinates" function.

```{r Cleaning - Part 2} 

sea_coords <- function(df){
  coords_land <- clean_coordinates(df, lon = "decimalLongitude", lat = "decimalLatitude", species = "species", tests = c("seas"), value="spatialvalid")
  filter(coords_land,.sea==FALSE) %>%
   select(species,decimalLongitude,decimalLatitude)
}

cleaned_list<-list()
for(i in 1:length(filtered_list)){
  cleaned_list[[i]]<-sea_coords(data.frame(filtered_list[[i]]))
}

```

Dataframes are often more easy to handle compared to lists. Therefore, the following function transforms lists into one big dataframe.

```{r}
list_to_df <- function(data_list) {
  comb_df <- data_list[[1]]
   for (l in 2:length(data_list)) {
     comb_df <- bind_rows(comb_df,data.frame(data_list[[l]]))
   }
  return(comb_df)
}
```

To get an overview of the distribution and to identify potential diversity hotspots of the different orders, the next step consists in the generation of global species richness plots for each respective order.

```{r Raster & Mapping}

filtered_df <- list_to_df(filtered_list); # might not need this!!!
cleaned_coordinates <- sea_coords(filtered_df); # do I really need this????


map_coral <- function(list_clean){
  for (l in 1:length(list_clean)) {
    corpoints <- st_as_sf(data.frame(list_clean[[l]]), coords = c(x="decimalLongitude", y="decimalLatitude"))
     r <- raster(corpoints, res = 1.5)
     r <- rasterize(corpoints, r, field = "species", fun= function(x, ...) {length(unique(na.omit(x)))})
     pdf(paste0(l,".pdf"))
     maps::map("world")
     plot(r, col = plasma(15), add=TRUE)  # adds the raster to the plot of the worldmap ##### breaks=cuts,
     dev.off()
   }
}

map_coral(cleaned_list);

```

The above code generates a map of global diversity richness for each order. Given the huge amount of data, the information displayed on these maps may appear a little overwhelming, and allow for little comparison between the orders. Therefore, a new map representing only the true hotspots of each respective order will be created.

```{r}

hotspots_order_list <- function(list_clean){
  list_hotspot <- list();
  
  for (l in 1:length(list_clean)){
     corpoints <- st_as_sf(data.frame(cleaned_list[[l]]), coords = c(x="decimalLongitude", y="decimalLatitude"))
     r <- raster(corpoints, res = 1.5)
     r <- rasterize(corpoints, r, field = "species", fun= function(x, ...) {length(unique(na.omit(x)))})
     hotspots <- data.frame(rasterToPoints(r, fun= function(x){x> 0.75*maxValue(r)}), Ordernumber=l)
     hotspots$percentage<-hotspots$layer*100/max(hotspots$layer,na.rm=TRUE)
     list_hotspot[[l]] <- hotspots
  }
  return(list_hotspot);
}

hotspot_list <- hotspots_order_list(cleaned_list);

df<-bind_rows(hotspot_list)

coral_names<-data.frame(Ordernumber=1:5,names=c("Alcyonacea","Corallimorpharia","Helioporacea","Scleractinia","Zoantharia"))

df<-left_join(df,coral_names)

df$Ordernumber<-as.character(df$Ordernumber)

world<-ne_countries(scale="medium",returnclass="sf")
div_col <- plasma(15)
ggplot(data=world)+geom_sf()+geom_point(data= df, aes(x=x,y=y,color=percentage,shape=names), size = 2) +scale_colour_gradientn(colours = div_col[1:12]) + xlab("Longitude (°)") + ylab("Latitude (°)") + ylim(-50,60) + xlim(-170,170) + labs(title = "Species richness hotspots per order") + theme(plot.title = element_text(hjust = 0.5), axis.text = element_text(size=12), legend.text = element_text(size=8), legend.key.size = unit(0.4, "cm"))


```

Coral reefs are complex systems that constitute an interspecific coral community that extends throughout different orders. Therefore, the next step aims to visualize the overal coral diversity hotspots of all orders combined on one map.

```{r Hotspot map of all coral species}

combine_all_coords <- function(list_all_orders) {
  c <-1
  coords_list <- list()
    for (l in 1:length(list_all_orders)){
       coords_list[[c]] <- st_as_sf((data.frame(list_all_orders[[l]])), coords = c(x="decimalLongitude", y="decimalLatitude"))
       c = c+1
    }  
  return(coords_list)
}

coords_all_sp <- list();
coords_all_sp <- combine_all_coords(cleaned_list);

df_all_coords <- coords_all_sp[[1]]
df_all_coords <- do.call('rbind',coords_all_sp);

r_all_sp <- raster(coords_all_sp[[1]], res= 1.5)
r_all_sp <- rasterize(df_all_coords, r_all_sp, field = "species", fun= function(x, ...) {length(unique(na.omit(x)))})
maps::map("world")
plot(r_all_sp, col = plasma(15), add=TRUE)

```

To highlight the coral diversity hotspots, let's only consider the regions that contain more than 200 species.

```{r}

hotspots_all_sp <-data.frame(rasterToPoints(r_all_sp,fun=function(x){x>200}))

world<-ne_countries(scale="medium",returnclass="sf")
div_col <- plasma(15)
ggplot(data=world)+geom_sf()+geom_point(data= hotspots_all_sp, aes(x=x,y=y,color=layer)) +scale_colour_gradientn(colours = div_col[1:12]) + xlab("Longitude (°)") + ylab("Latitude (°)") + ylim(-50,60) + xlim(-170,170) + labs(title = "Species richness hotspots of the world") + theme(plot.title = element_text(hjust = 0.5), axis.text = element_text(size=12))
```

Coral bleaching susceptibility vs resistance

Different species of corals have different levels of tolerance with respect to fluctuating temperatures. Swain et al. (2016) have defined and standardized this variability in the form of a "Bleaching Response Index" (BRI). In total, they have defined BRIs for 

The data has been downloaded from the supplementary material provided by Swain et al. (2016). 

```{r BRI data}

BRI_data <- fread("BRI_taxon_Swain2016.csv", select = c("Taxon", "Taxon-BRI (Bleaching Response Index) (%)", "Number of records", "Std Deviation"))

BRI_data$`Taxon-BRI (Bleaching Response Index) (%)`<-as.numeric(gsub(",",".",BRI_data$`Taxon-BRI (Bleaching Response Index) (%)`))

BRI_data$`Std Deviation`<-as.numeric(gsub(",",".",BRI_data$`Std Deviation`))

clean_BRI_data <- filter(BRI_data, BRI_data$`Number of records` != "1" & BRI_data$`Std Deviation` != "") # remove records where standard deviation is zero and those where the number of records is only one


# filter out genus (character length = 1)
species_only <- function(taxa) {
  split_taxa <- list()
  length_taxa <- integer()
  for (i in 1:length(taxa)) {
    split_taxa <- strsplit(taxa[i]," ")
    length_taxa <- length(split_taxa[[1]])
      if (length_taxa == 1) {
        taxa[i] <- NA
      }
  }
  return(taxa)
}

clean_BRI_data$Taxon <- species_only(clean_BRI_data$Taxon);
clean_BRI_data <- na.omit(clean_BRI_data)
```

Standardized taxon-specifc Bleaching Response Index (BRIj) defined by Swain et al. (2016):
- Highly susceptible corals have a BRIj above 40%
- Highly resistant corals have a BRIj below 10%

```{r}
susceptible_corals <- clean_BRI_data %>% filter(`Taxon-BRI (Bleaching Response Index) (%)` > 40);
resistant_corals <- clean_BRI_data %>% filter(`Taxon-BRI (Bleaching Response Index) (%)` < 10);
```

Function that finds susceptible/resistant species in GBIF data set.

```{r}

GBIF_susc <- filter(df_all_coords, df_all_coords$species %in% susceptible_corals$Taxon);
GBIF_res <- filter(df_all_coords, df_all_coords$species %in% resistant_corals$Taxon);

```

Mapping interspecific susceptibility

```{r}

r_susc <- r_all_sp
reset_vector <- integer(length(ncell(r_susc)));
r_susc <- setValues(r_susc,reset_vector);
r_susc <- rasterize(GBIF_susc, r_susc, field = "species", fun= function(x, ...) {length(unique(na.omit(x)))})

pts_susc<-data.frame(rasterToPoints(r_susc,fun=function(x){x>10}))
world<-ne_countries(scale="medium",returnclass="sf")
ggplot(data=world)+geom_sf()+geom_point(data=pts_susc,aes(x=x,y=y,color=layer))+scale_colour_gradientn(colours = c("#FFCCCC", "#FF9999","#FF6666","#CC3333", "#990000", "#660000")) + xlab("Longitude (°)") + ylab("Latitude (°)") + ylim(-50,60) + labs(title = "Interspecific susceptibility hotspots of the world") + theme(plot.title = element_text(hjust = 0.5)) 
```

Coral susceptibility with respect to species richness (i.e. coral diversity)

```{r}
r_all_new<-r_all_sp>10
r_all_newer<-r_all_sp*r_all_new

r_all_newer<-setValues(r_all_newer,ifelse(getValues(r_all_sp)>10,
                             getValues(r_all_sp),
                             NA))

  
r_rel_susc <- r_susc/r_all_newer;

maps::map("world")
plot(r_rel_susc, add= TRUE)

sus.df<-data.frame(num.sp=getValues(r_susc),num.obs=getValues(r_susc))
ggplot(sus.df,aes(x=num.obs,y=num.sp))+geom_point()

r_rel_susc_high <- r_rel_susc > 0.2  
pts_high_susc <- data.frame(rasterToPoints(r_rel_susc,fun=function(x){x>0.2}))

world<-ne_countries(scale="medium",returnclass="sf")
ggplot(data=world)+geom_sf()+geom_point(data=pts_high_susc,aes(x=x,y=y,color=layer))+scale_colour_gradientn(colours = c("#FFCCCC", "#FF9999","#FF6666","#CC3333", "#990000", "#660000")) + xlab("Longitude (°)") + ylab("Latitude (°)") + ylim(-50,60) + labs(title = "Interspecific susceptibility hotspots of the world with respect to coral diversity") + theme(plot.title = element_text(hjust = 0.5))

```
Mapping interspecific resistance

```{r}
r_res <- r_all_sp
reset_vector_res <- integer(length(ncell(r_res)));
r_res <- setValues(r_res,reset_vector_res);
r_res <- rasterize(GBIF_res, r_res, field = "species", fun= function(x, ...) {length(unique(na.omit(x)))})

pts_res<-data.frame(rasterToPoints(r_res,fun=function(x){x>3}))
world<-ne_countries(scale="medium",returnclass="sf")
col_res <- brewer.pal(8,"Blues")
ggplot(data=world)+geom_sf()+geom_point(data=pts_res,aes(x=x,y=y,color=layer))+scale_colour_gradientn(colours = col_res[4:8]) + xlab("Longitude (°)") + ylab("Latitude (°)") + ylim(-50,60) + labs(title = "Interspecific resistance 'hotspots' of the world") + theme(plot.title = element_text(hjust = 0.5)) 
```

Interspecific resistance with respect to coral diversity

```{r}

r_rel_res <- r_res/r_all_newer;

maps::map("world")
plot(r_rel_res, add= TRUE)

res.df<-data.frame(num.sp=getValues(r_res),num.obs=getValues(r_res))
ggplot(res.df,aes(x=num.obs,y=num.sp))+geom_point()

pts_high_res <- data.frame(rasterToPoints(r_rel_res,fun=function(x){x>0.2}))

world<-ne_countries(scale="medium",returnclass="sf")
ggplot(data=world)+geom_sf()+geom_point(data=pts_high_res,aes(x=x,y=y,color=layer))+scale_colour_gradientn(colours = col_res[4:8]) + xlab("Longitude (°)") + ylab("Latitude (°)") + ylim(-50,60) + labs(title = "Interspecific resistance hotspots of the world with respect to coral diversity") + theme(plot.title = element_text(hjust = 0.5))

```

Susceptibility of respective orders

```{r}

# Count of number of species in the different orders
species_in_Alcyonacea <- unique.data.frame(data.frame(cleaned_list[[1]]$species));
species_in_Corallimorpharia <- unique.data.frame(data.frame(cleaned_list[[2]]$species));
species_in_Helioporacea <- unique.data.frame(data.frame(cleaned_list[[3]]$species));
species_in_Scleractinia <- unique.data.frame(data.frame(cleaned_list[[4]]$species));
species_in_Zoantharia <- unique.data.frame(data.frame(cleaned_list[[5]]$species));

susc_Alcy <- filter(species_in_Alcyonacea, species_in_Alcyonacea[[1]] %in% susceptible_corals$Taxon);
susc_Scler <- filter(species_in_Scleractinia, species_in_Scleractinia[[1]] %in% susceptible_corals$Taxon);
res_Alcy <- filter(species_in_Alcyonacea, species_in_Alcyonacea[[1]] %in% resistant_corals$Taxon);
res_Scler <- filter(species_in_Scleractinia, species_in_Scleractinia[[1]] %in% resistant_corals$Taxon);

res_Cor <- filter(species_in_Corallimorpharia, species_in_Corallimorpharia[[1]] %in% resistant_corals$Taxon);
res_Helio <- filter(species_in_Helioporacea, species_in_Helioporacea[[1]] %in% resistant_corals$Taxon);

```

Bleaching sites data from Donner et al. (2017): Worldwide recorded coral bleaching from 1963-2011

```{r}
Bleaching_data <- fread("Data_Donner_2017.csv", select = c("LATITUDE", "LONGITUDE", "YEAR", "SEVERITY_CODE"))
clean_bleaching_data <- filter(Bleaching_data, (Bleaching_data$SEVERITY_CODE == "2" | Bleaching_data$SEVERITY_CODE == "3"))
clean_bleaching_data$LATITUDE <- as.numeric(gsub(",",".", clean_bleaching_data$LATITUDE))
clean_bleaching_data$LONGITUDE <- as.numeric(gsub(",",".", clean_bleaching_data$LONGITUDE))

```

Animation of bleached sites from 1973 to 2011

```{r}
world <- ne_countries(scale="medium",returnclass="sf")

#rasterToPoints(r_susc,fun=function(x){x>10}))

bleach_map <- function(year){
  
    ggplot(data=world) + geom_sf() + geom_point(data= clean_bleaching_data[clean_bleaching_data$YEAR == year,],
                        aes(x=LONGITUDE, y=LATITUDE, color = SEVERITY_CODE)) +
                        theme_bw() + ylab("Latitude") + xlab("Longitude") + xlim(-180,180) + ylim(-90,90)
    print(paste0("saving plot ", year))
    ggsave(filename = paste0("bleach_events_in_",year,".png"), width = length(xlim), height = length(ylim))
}


seq(from = 1973, to=2011, by=1) %>% 
  map_df(bleach_map)

list.files(path = "bleach_events_in", pattern = "*.png") %>%
  map(image_read) %>% # reads each path file
  image_join() %>% # joins image
  image_animate(fps=2) %>% # animates 2 frames per second
  image_write("bleaching_events.gif") # write to current dir
```

Bleach-susceptible regions in the period from 1973 to 2011

```{r}

cor_bleach <- st_as_sf(clean_bleaching_data, coords = c(x="LONGITUDE", y="LATITUDE"))
r_bleach <- raster(cor_bleach, res = 1.5)
#r_bleach <- rasterize(cor_bleach, r_bleach)
r_bleach <- rasterize(cor_bleach, r_res, field = "YEAR", fun= function(x, ...) {length(na.omit(x))})
maps::map("world")
plot(r_res, add=TRUE)

```

Hotspots of bleaching events from 1973-2011

```{r}
pts_bleach<-data.frame(rasterToPoints(r_bleach,fun=function(x){x>10}))
world<-ne_countries(scale="medium",returnclass="sf")
col_res <- brewer.pal(8,"Greys")
ggplot(data=world)+geom_sf()+geom_point(data=pts_bleach,aes(x=x,y=y,colour = layer))+scale_colour_gradientn(colors = col_res[3:8]) + xlab("Longitude (°)") + ylab("Latitude (°)") + ylim(-50,60) + labs(title = "Bleaching hotspots worldwide from 1973 to 2011") + theme(plot.title = element_text(hjust = 0.5)) 
```

